Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-30 paper_nips/paper.tex
update

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\DeclareMathOperator{\proj}{proj}
\DeclareMathOperator{\soft}{soft}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator{\Prox}{Prox}
\DeclareMathOperator{\im}{im}

@@

\begin{equation}
  \mathbb E\left[{\mathcal{L_{AE}}}(\mathbf{x})\right] = %
  \mathbb E\left[\| \mathbf{x} - f_{AE}(\mathbf{x}) \|^2\right],
\end{equation}
with \(f_{AE}(\mathbf x) = %
\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
%% This reconstruction error criterion equates with
%% maximizing a lower bound on the mutual information between
%% input and the learned representation.
\suggestadd{
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$. In particular, this means that the weights
learned are forced to take a two-fold function: That of signal 
\textit{analysis} and that of signal \textit{synthesis}.
Whereas the first layer \textit{analyzes}
the data in order to obtain the cleanest latent representation, the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations. Tying these functionalities makes the analysis
side interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.
}
\paragraph{Reduced-rank logistic regression.}
Lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the low-rank
multinomial logistic regression.
It gives the probability of an input $\mathbf{x}$ to belong
to a class $i \in \mathbb{N}$

\begin{eqnarray}
  \begin{split}
    P(Y=i|\mathbf{x, V_0,V_1,c_0, c_1}) &= softmax_i(\mathbf{V_1 (V_0 x + c_0) + c_1}) \\
    &= \frac {e^{\mathbf{V_{1i} (V_{0i} x + b_{V_{0i}}) + b_{V_{1i}}}}} {\sum_j e^{\mathbf{V_{1j} (V_{0j} x + b_{V_{0j}}) + b_{V_{1j}}}}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

where the matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxc}}$
projects the latent components
onto hyperplanes that reflect $c$ class probabilities.
$\mathbf{c_0}$ and $\mathbf{c_1}$ are corresponding
bias vectors.
The loss function is given by

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(\mathbf{x, y}) = \frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|\mathbf{x^{(i)}; V_1 (V_0 x + c_0) + c_1})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the low-rank logistic regression
are linked one two levels. First, their transformation matrices mapping from
input to the latent space are identical
\begin{eqnarray}
  \mathbf{V_0} = \mathbf{W_0}
\end{eqnarray}
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent an optimal latent code for both
rest and task activity data.
Second, the objectives of the autoencoder and the low-rank
logistic regression are interpolated in the common loss function

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = \lambda {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \ell_1(\theta) + \ell_2(\theta)
  \label{eq:loss_equ}
\end{split}
\end{eqnarray}

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,c_0, c_1, b_0, b_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
For parameter shrinkage, we
impose a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$ (i.e., elasticnet).

\paragraph{Optimization.}
The common objective was approximated 
% XXX why 'approximated' ?
in the neuroimaging data
by updating parameters by derivates 
% XXX derivates ???
of the 
semi-supervised low-rank logistic regression.
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives through the linear network.
As solver, we chose \textit{rmsprop} \cite{rmsprop},
a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by rmsprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to zero.
With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by

>>>>

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\DeclareMathOperator{\proj}{proj}
\DeclareMathOperator{\softmax}{softmax}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator{\Prox}{Prox}
\DeclareMathOperator{\im}{im}

@@

\begin{equation}
  \mathbb E\left[{\mathcal{L_{AE}}}(\mathbf{x})\right] = %
  \mathbb E\left[\| \mathbf{x} - f_{AE}(\mathbf{x}) \|^2\right],
\end{equation}
with \(f_{AE}(\mathbf x) = %
\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
%% This reconstruction error criterion equates with
%% maximizing a lower bound on the mutual information between
%% input and the learned representation.
\suggestadd{
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$. In particular, this means that the weights
learned are forced to take a two-fold function: That of signal 
\textit{analysis} and that of signal \textit{synthesis}.
Whereas the first layer \textit{analyzes}
the data in order to obtain the cleanest latent representation, the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations. Tying these functionalities makes the analysis
side interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.
}
\paragraph{Reduced-rank logistic regression.}
Lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the low-rank
multinomial logistic regression.
It gives the probability of an input $\mathbf{x}$ to belong
to a class $i \in \{1, \dots, l\}$

\begin{equation}
  \begin{split}
    P(Y=i|\mathbf{x; V_0,V_1,c_0, c_1}) &= \softmax_i(f_{RL}(\mathbf x))
  \end{split}
  \label{eq:lr}
\end{equation}
where \(f_{LR}(\mathbf x) = \mathbf{V_1 (V_0 x + c_0) + c_1}\) computes 
multinomial logits and \(\softmax_i(x) = \exp(x_i)/\sum_j\exp(x_j)\).
The matrix $\mathbf{V_0 \in \mathbb{R}^{dxn}}$
transforms the input $\mathbf{x \in \mathbb{R}^{d}}$
into $n$ latent components
and the matrix $\mathbf{V_1 \in \mathbb{R}^{nxl}}$
projects the latent components
onto hyperplanes that reflect $l$ label probabilities.
$\mathbf{c_0}$ and $\mathbf{c_1}$ are corresponding
bias vectors.
The loss function is given by

\begin{equation}
    \mathbb E\left[{\mathcal{L_{LR}}}(\mathbf{x, y})\right] \approx %
\frac{1}{N_{X_{task}}} \sum_{k=0}^{N_{X_{task}}} \log(P(Y=y^{(k)}|\mathbf{x^{(k)}; V_0, V_1, c_0, c_1})
\label{eq:lr_loss}
\end{equation}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the low-rank logistic regression
are linked on two levels. First, their transformation matrices mapping from
input to the latent space are tied
\begin{eqnarray}
  \mathbf{V_0} = \mathbf{W_0}.
\end{eqnarray}
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent an optimal latent code for both
rest and task activity data.
Second, the objectives of the autoencoder and the low-rank
logistic regression are interpolated in the common loss function

% XXX Michael: In resolving conflict I opted for my change. Feel free to revert if that was intended.
\begin{equation}
{\mathcal{L}}(\theta, \lambda) = \lambda {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \Omega(\theta).
  \label{eq:loss_equ}
\end{equation}

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,c_0, c_1, b_0, b_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
\(\Omega(\theta)\) represents a regularization, for which we choose the
ElasticNet type, i.e. a combination of $\ell_1$ and $\ell_2$ penalties
$\forall p \in \theta$.

% XXX Michael: To sum up the math in this paragraph: We need to decide what
% XXX the \mathcal L are supposed to mean - sample wise loss or expected loss
% XXX or some global deterministic loss (the latter would be bad, because we)
% XXX are not optimizing it. Right now it looks like a sample-wise loss.

\paragraph{Optimization.}
The common objective was \suggestremove{approximated} \suggestadd{optimized}
% XXX why 'approximated' ?
in the neuroimaging data
by \suggestremove{updating parameters by derivates}
\suggestadd{gradient descent in the parameters}
% XXX derivates ???
of the 
semi-supervised low-rank logistic regression.
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives through the linear network.
As solver, we chose \textit{rmsprop} \cite{rmsprop},
a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by rmsprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to zero.
With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by
